Finch: Sparse and Structured Array Programming with Control Flow

### 2024-03-21 sections/lang.tex
more rules ugh

\[
    \frac{}{\langle\finchliteral(val), (e, t)\rangle \rightarrow val} \quad
    \frac{}{\langle\finchvalue(ex, type), (e, t)\rangle \Rightarrow \langle ex, t \rangle} \quad
\]
\[
    \frac{}{\langle\finchvar(name), (e, t)\rangle \rightarrow e(\finchvar(name))} \quad
    \frac{}{\langle\finchindex(name), (e, t)\rangle \rightarrow e(\finchindex(name))}
\]
\[
    \frac{\langle args_i, (e, t) \rangle \Rightarrow vals_i \quad \langle f, (e, t) \rangle \Rightarrow g}
    {\langle\finchcall(f, args...), (e, t)\rangle \rightarrow \langle g(vals...), t \rangle}
\]
\[
\frac{\langle head, s_1 \rangle \rightarrow s_2}
{\langle\finchblock(head, tail...), s_1 \rangle \rightarrow \langle \finchblock(tail...), s_2 \rangle} \quad
\frac{}{\langle\finchblock(), s\rangle \rightarrow s}
\]
\[
\frac{\langle cond, s \rangle \Rightarrow true}
{\langle\finchsieve(cond, body), s\rangle \rightarrow \langle body, s \rangle}\quad
\frac{\langle cond, s \rangle \Rightarrow false}
{\langle\finchsieve(cond, body), s\rangle \rightarrow s}
\]
\[
\frac{\langle val, (e, t) \rangle \rightarrow val'}
{\langle\finchdefine(var, val, body), (e, t)\rangle \rightarrow \langle body, (e[var \mapsto val'], t) \rangle}
\]
\[
\frac{s = (e, t) \quad e(\finchmode(tns)) = read \quad e(tns) = tns' \quad  \langle init, s \rangle \rightarrow init' \quad  \langle dims_i, s \rangle \rightarrow dims_i'}
{\langle\finchdeclare(tns, init, dims...), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \langle declare(tns', init', dims'...), t \rangle)}
\]
\[
\frac{s = (e, t) \quad e(\finchmode(tns)) = read \quad e(tns) = tns'}
{\langle\finchthaw(tns), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \langle thaw(tns'), t \rangle)}
\]
\[
\frac{s = (e, t) \quad e(\finchmode(tns)) = update \quad e(tns) = tns'}
{\langle\finchfreeze(tns), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto read], \langle freeze(tns'), t \rangle)}
\]

$E$ is an evaluation context that applies anywhere in the syntax tree.

\[
\frac{e(tns) = tns' \langle unfurl(tns'), t\rangle \rightarrow \langle tns'', t'\rangle}
{\langle \finchloop(i, ext, E[\finchaccess(tns, j..., i)]), (e, t)\rangle \rightarrow \langle \finchloop(i, ext, E[\finchaccess(tns'', j..., i)]), (e, t')\rangle}
\]

\[
\frac{}
{\splitfrac{\langle \finchloop(i, ext, E[\finchaccess(sequence(phase(range, body), tail...), j..., i)]), s\rangle
    }{\rightarrow \langle \finchblock(\finchloop(i, ext \cup range, E[\finchaccess(body, j..., i)]), \finchloop(i, ext, E[\finchaccess(sequence(tail...), j..., i)])), s\rangle}}
\]

\[
\frac{}
{\langle \finchloop(i, ext, E[\finchaccess(sequence(), j..., i)]), s\rangle \rightarrow \langle \finchblock(), s\rangle}
\]

\[
\frac{}
{\splitfrac{\langle \finchloop(i, extent(a, b), E[\finchaccess(stepper(next, body), j..., i)]), s\rangle
    }{\rightarrow \langle \finchblock(\finchloop(i, ext \cup range, E[\finchaccess(body, j..., i)]), \finchloop(i, ext, E[\finchaccess(sequence(tail...), j..., i)])), s\rangle}}
\]

>>>>

the state $s$ of the finch compiler is a tuple $(e, t)$ of a variable value
environment and another state $t$ corresponding to the embedding language.

\begin{figure}
    \centering
    \small

    \begin{prooftree}  
    \hypo{s = (e, t)}
    \hypo{e(\finchmode(tns)) = read}
    \hypo{e(tns) = tns'}
    \infer3[$Thaw$]{\langle\finchthaw(tns), s\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \langle thaw(tns'), t \rangle)}
    \end{prooftree}
    \vspace{12pt}

    \begin{prooftree}  
    \hypo{s = (e, t)}
    \hypo{e(\finchmode(tns)) = update}
    \hypo{e(tns) = tns'}
    \infer3[$Freeze$]{\langle\finchfreeze(tns), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto read], \langle freeze(tns'), t \rangle)}
    \end{prooftree}
      \vspace{12pt}

    \begin{prooftree}  
    \hypo{s = (e, t)}
    \hypo{e(tns) = tns'}
    \hypo{\langle init, s \rangle \Rightarrow init'}
    \hypo{\forall i \langle init, dims_i \rangle \Rightarrow dims'_i}
    \infer4[$Declare$]{\langle\finchdeclare(tns, init, dims), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \langle \finchdeclare(tns', init', dims'...), t \rangle)}
    \end{prooftree}

    \caption{Rules for evaluating life cycle functions.}
    \label{fig:enter-label}
\end{figure}

\begin{figure}
    \centering
    \small

    \begin{prooftree}
    \hypo{\langle val, (e, t) \rangle
    \rightarrow val'}
    \infer1[$Define$]{\splitfrac{\langle\finchdefine(var, val, body), (e, t)\rangle}{\rightarrow \langle body, (e[var \mapsto val'], t) \rangle}}
    \end{prooftree}
    \hfill
    \begin{prooftree}
        \hypo{\langle args_i, (e, t) \rangle \Rightarrow vals_i}
        \hypo{\langle f, (e, t) \rangle \Rightarrow g}
        \infer2[$Call$]{\langle\finchcall(f, args...), (e, t)\rangle \rightarrow \langle g(vals...), t \rangle}
    \end{prooftree}
    \vspace{12pt}

    \begin{prooftree}
        \hypo{}
        \infer1[$Literal$]{\langle\finchliteral(val), (e, t)\rangle \rightarrow val}
    \end{prooftree}
    \hfill
    \begin{prooftree}
        \hypo{}
        \infer1[$Value$]{\langle\finchvalue(ex, type), (e, t)\rangle \Rightarrow \langle ex, t \rangle}
    \end{prooftree}
    \vspace{12pt}

    \begin{prooftree}
        \infer0[$Variable$]{\splitfrac{\langle\finchvar(name), (e, t)\rangle}{\rightarrow e(\finchvar(name))}}
    \end{prooftree}
    \hfill
    \begin{prooftree}
        \infer0[$Index$]{\splitfrac{\langle\finchindex(name), (e, t)\rangle}{\rightarrow e(\finchindex(name))}}
    \end{prooftree}
    \vspace{12pt}

    \begin{prooftree}
        \hypo{\langle head, s_1 \rangle \rightarrow s_2}
        \infer1[$BlockHead$]{\langle\finchblock(head, tail...), s_1 \rangle \rightarrow \langle \finchblock(tail...), s_2 \rangle}
    \end{prooftree}
    \hfill
    \begin{prooftree}
        \infer0[$EmptyBlock$]{\langle\finchblock(), s\rangle \rightarrow s}
    \end{prooftree}
    \vspace{12pt}

    \begin{prooftree}
        \hypo{\langle cond, s \rangle \Rightarrow true}
        \infer1[$SieveTrue$]{\langle\finchsieve(cond, body), s\rangle \rightarrow \langle body, s \rangle}
    \end{prooftree}
    \hfill
    \begin{prooftree}
        \hypo{\langle cond, s \rangle \Rightarrow false}
        \infer1[$SieveFalse$]{\langle\finchsieve(cond, body), s\rangle \rightarrow s}
    \end{prooftree}

    \caption{Basic Evaluation Rules}
    \label{fig:enter-label}
\end{figure}

\begin{figure}
    \centering
    \small

$E$ is an evaluation context that applies anywhere in the syntax tree.
\vspace{12pt}

\begin{prooftree}
    \infer0[$Run$]{\splitfrac{\langle \finchloop(i, ext, E[\finchaccess(run(body), j..., i)]), s\rangle}{\rightarrow \langle \finchloop(i, ext, E[\finchaccess(body, j...)]), s\rangle}}
\end{prooftree}
\hfill
\begin{prooftree}
    \hypo{e(i) = i'}
    \hypo{\langle seek(i'), t \rangle \rightarrow t'}
    \infer2[$Lookup$]{\splitfrac{\langle E[\finchaccess(lookup(seek, body), j..., i)], (e, t)\rangle}{\rightarrow \langle E[\finchaccess(body, j...)], (e, t')\rangle}}
\end{prooftree}
\vspace{12pt}

\begin{prooftree}
    \hypo{\langle cond, t \rangle \Rightarrow true}
    \infer1[]{\splitfrac{\langle E[\finchaccess(switch(cond, head, tail), i...)], s\rangle}{\rightarrow \langle E[\finchaccess(head, i...)], s\rangle}}
\end{prooftree}
\hfill
\begin{prooftree}
    \hypo{\langle cond, t \rangle \Rightarrow false}
    \infer1[$Switch$]{\splitfrac{\langle E[\finchaccess(switch(cond, head, tail), i...)], s\rangle}{\rightarrow \langle E[\finchaccess(tail, i...)], s\rangle}}
\end{prooftree}
\vspace{12pt}

\begin{prooftree}
    \infer0[$Phase$]{\splitfrac{\langle \finchloop(i, extent(a, b), E[\finchaccess(phase(extent(c, d), body), j..., i)]), s\rangle}{\rightarrow \langle \text{\finchloop}(i, extent(max(a, c), min(b, d)), E[\text{\finchaccess}(body, j..., i)]), s\rangle}}
\end{prooftree}
\vspace{12pt}

\begin{prooftree}
    \hypo{\langle \finchloop(i, ext, E[\finchaccess(head, j..., i)]), s\rangle \rightarrow s'}
    \infer1[$Sequence$]{\splitfrac{\langle \finchloop(i, ext, E[\finchaccess(sequence(head, tail), j..., i)]), s\rangle}{\rightarrow \langle \finchloop(i, ext, E[\finchaccess(tail, j..., i)]), s'\rangle}}
\end{prooftree}
\vspace{12pt}

\begin{prooftree}
    \hypo{\langle seek(a), t \rangle \rightarrow t'}
    \infer1[$StepperSeek$]{\splitfrac{\langle \finchloop(i, extent(a, b), E[\finchaccess(stepper(seek, body, next), j..., i)]), (e, t)\rangle}{\rightarrow \langle \finchloop(i, extent(a, b), E[\finchaccess(stepper(body, next), j..., i)]), (e, t')\rangle}}
\end{prooftree}
\vspace{12pt}

\begin{prooftree}
    \hypo{\langle \finchloop(i, ext, E[\finchaccess(body, j..., i)]), (e, t) \rangle \rightarrow (e', t')}
    \hypo{\langle next, t' \rangle \rightarrow t''}
    \infer2[$StepperNext$]{\splitfrac{\langle \finchloop(i, ext, E[\finchaccess(stepper(body, next), j..., i)]), s\rangle}{\rightarrow \langle \finchloop(i, ext, E[\finchaccess(stepper(body, next), j..., i)]), (e', t'')\rangle}}
\end{prooftree}

    \caption{Looplet Rules}
    \label{fig:enter-label}
\end{figure}
